Defining properties regulating pioneer-factor binding and function

### 2023-01-12 manuscript/manuscript_sections/_03_Results.qmd
add code for supp figs 2-4
add code for supp fig6
add code for supp fig7

Previous work has shown that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (refs).
<!--# Add a few sentences here with more details on past work --> To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
Analysis of average MNase-seq signal at class I-III peaks indicated that class III sites have high levels of nucleosome occupancy that, on average, occur over the respective motif.
Hierarchical clustering of MNase-seq data revealed that there was no preference for binding or chromatin opening at sites that had motifs in a particular orientation on nucleosomes (Extended Data Fig. XX).
For class II and III peaks, some motifs were positioned close to the nuclesome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
This analysis is complicated by the fact that MNase-seq is a bulk assay and may not accurately represent nucleosome positioning in individual cells, especially at regions with poorly positioned nucleosomes.
To more directly assess the positioning of nucleosomes around motifs, we analyzed a previously published single-molecule dataset using methyltransferase footprinting to determine the locations of nucleosomes on individual chromatin fibers [@Stergachis2020].
We analyzed the distance from motifs at class II and regions to the nearest nucleosome dyad.
This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
However, the lack of strong nucleosome positioning at these regions makes it difficult to fully assess the importance of motif positioning.

@@

Prior work has shown that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs).
However, little is known about how regions outside the DBD may contribute to pioneering function.
<!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat. Genet. paper) is outside the DBD. See also new Zaret paper in NSMB looking at contribution of IDRs and activation domains to chromatin opening on nucleosome arrays in vitro -->.
To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig. XX).
Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was totally unable to bind to class II or III regions (Fig. 6a,c).
We had previously shown that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
Because many Zld binding sites in S2 cells also lack a canonical CAGGTA motif (Fig. 3, Extended Data Fig XX), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required.
Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
Together, these experiments show that regions outside the DBD are essential for binding and opening of closed chromatin.

>>>>

Previous work has shown that binding of some pioneer factors depends on how target motifs are positioned on nucleosomes (refs).
<!--# Add a few sentences here with more details on past work --> To test if pioneering by Zld, Grh or Twi was associated with motif positioning on nucleosomes, we analyzed published MNase-seq data from WT S2 cells [@Chereji2019].
Analysis of average MNase-seq signal at class I-III peaks indicated that class III sites have high levels of nucleosome occupancy that, on average, occur over the respective motif.
Hierarchical clustering of MNase-seq data revealed that there was no preference for binding or chromatin opening at sites that had motifs in a particular orientation on nucleosomes (Extended Data Fig. XX).
For class II and III peaks, some motifs were positioned close to the nuclesome dyad, some motifs were at the edge of nucleosomes, and some motifs were in linker DNA.
This analysis is complicated by the fact that MNase-seq is a bulk assay and may not accurately represent nucleosome positioning in individual cells, especially at regions with poorly positioned nucleosomes.
To more directly assess the positioning of nucleosomes around motifs, we analyzed a previously published single-molecule dataset using methyltransferase footprinting to determine the locations of nucleosomes on individual chromatin fibers [@Stergachis2020].
We analyzed the distance from motifs at class II and III regions to the nearest nucleosome dyad.
This analysis showed that nucleosomes at class II and III regions are poorly positioned (Extended Data Fig. XX).
Comparing class II and III regions did not identify any differences in nucleosome positioning that might be associated with chromatin opening at class III regions (Extended Data Fig. XX).
Together, these data suggest that Zld, Grh and Twi do not have a strong preference for motifs at a particular position on nucleosomes.
However, the lack of strong nucleosome positioning at these regions makes it difficult to fully assess the importance of motif positioning.

@@

Prior work has shown that binding to nucleosomes by pioneer factors requires recognition of target motifs by DNA-binding domains (DBDs) (REFs).
However, little is known about how regions outside the DBD may contribute to pioneering function.
<!--# Consider toning down this statement, as the FoxA1 region that interacts with nucleosomes (Iwafuchi 2020 Nat. Genet. paper) is outside the DBD. See also new Zaret paper in NSMB looking at contribution of IDRs and activation domains to chromatin opening on nucleosome arrays in vitro -->.
To test if regions outside the DBD are required for pioneering activity, we performed ChIP-seq and ATAC-seq on stable cell lines expressing the Zld or Grh DBD at levels comparable to the full-length (FL) protein (Extended Data Fig. XX).
Analysis of ChIP-seq signal at our previously defined class I-III regions revealed that the Zld DBD bound robustly to class I regions, but was totally unable to bind to class II or III regions (Fig. 6a,c).
We had previously shown that in neural stem cells, Zld binds to many regions that lack a canonical Zld motif through a mechanism that is not understood [@Larson2021a].
Because many Zld binding sites in S2 cells also lack a canonical CAGGTA motif (Fig. 3, Extended Data Fig XX), it is possible that expressing the DNA-binding domain alone also removed regions of the protein that are required for CAGGTA-independent binding.
To rule out this possibility, we analyzed DBD ChIP-seq signal only at regions containing a canonical motif.
This analysis demonstrated that the DBD alone is unable to bind to these regions, suggesting that even for binding to canonical motifs, regions outside the DBD are required (Extended Data Fig. XX).
Binding of the Grh DBD to class II-III regions was strongly reduced compared to the FL protein, although some closed regions retained low levels of binding (Fig. 6b,e).
For either Zld or Grh, the DBD alone was insufficient to open chromatin (Fig 6d,f).
Together, these experiments show that regions outside the DBD are essential for binding and opening of closed chromatin.

### 2023-01-12 manuscript/manuscript_sections/_06_Figure_Legends.qmd
add code for supp fig6
add code for supp fig7

>>>>

**Extended Data Figure 7: Zld, Grh and Twi display cell-type specific binding**

-   comparison to other tissues

-   analysis of motifs across tissues?

**Extended Data Figure 8: PRC2 inhibition is insufficient to promote Zld, Grh, and Twi binding to repressed targets**

-   Examples of class III regions with H3K27me3

-   examples of class IV-VII regions

-   Spike-in validation of H3K27me3 antibody

-   Genome browsers showing loss of H3K27me3 in taz-treated cells without changes in binding.

**Extended Data Figure 9: Expression of Zld, Grh, or Twi at supraphysiological levels results in novel binding sites**

-   Western blot showing protein levels for titration experiments

-   Examples of novel binding sites for each factor

-   motif enrichment at novel binding sites

-   Chromatin features at novel binding sites.

**Extended Data Figure 10:**

-   Analysis of Zld DBD ChIP only at CAGGTAG sites

-   Examples of any novel DBD binding sites not detected for full-length proteins
